---
title: "Measuring Political Ad Tone"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Measuring Political Ad Tone}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
library(textscale)
library(ggplot2)
library(dplyr)
```

Political scientists often need to measure latent quantities from text —
concepts like negativity, ideology, argument strength, etc. Carlson & Montgomery (2017) developed a pairwise
comparison approach to this problem which they call `SentimentIt`: instead of rating each document in isolation, annotators are shown pairs of documents and asked "Which one is more negative?" Aggregating these judgments with a scaling model yields a continuous score for every document.

The `textscale` package extends this approach by combining pairwise annotations with text
embeddings. Rather than requiring comparisons between every pair of documents,
textscale fits a ridge logistic regression on embedding differences to
identify a *direction* in embedding space that separates winners from losers.
Once that direction is learned, any document can be scored by projecting its
embedding onto it — including documents that were never directly compared.

## Data

Carlson & Montgomery (2017) published the pairwise annotations from their Wisconsin
congressional ads study alongside their paper. The `ad_tone` dataset bundled
with textscale contains 935 ads from that study, together with `textscale`
scores derived from those annotations.

```{r data-overview}
data(ad_tone)
glimpse(ad_tone)
```

## Scaling with `textscale`

To replicate this analysis from scratch — or to apply it to your own documents — you would run `textscale()` with a single prompt:

```{r textscale-call, eval = FALSE}
result <- textscale(
  documents = ad_tone$text,
  prompt    = "Which political ad is more negative toward its opponent?",
  seed      = 1847
)
```

This handles the full pipeline: generating comparison pairs, retrieving
embeddings, submitting annotations to an LLM via the OpenAI Batch API,
fitting and validating the model, and returning scores for all documents.
An OpenAI API key stored in the `OPENAI_API_KEY` environment variable is
required.

## Replicating with the crowd-sourced annotations

The `cm_comparisons` dataset contains the original crowd-coded pairwise comparisons
from Carlson & Montgomery (2017). You can use these to estimate `textscale` scores directly without any calls to the LLM.

```{r textscale-cached, eval = FALSE}
# 1. Get embeddings for each ad
emb <- get_embeddings(ad_tone$text)

# 2. Fit model on cm_comparisons (train set only by default)
mod <- fit_model(cm_comparisons, emb)

# 3. Check model calibration against held-out ads
validate_model(mod, cm_comparisons, emb)

# 4. Fit final model on all pairwise comparisons
mod <- fit_model(cm_comparisons, emb, refit = TRUE)

# 5. Project scores
scores <- score_documents(mod, emb, ci = TRUE)
```

## Validation

The `ad_tone_validation` object contains the results of evaluating the model
on a held-out 20% test split of documents.

```{r validation}
data(ad_tone_validation)
print(ad_tone_validation)
```

The model correctly predicts the more negative ad in about 80% of held-out
comparisons. The calibration plot confirms that predicted probabilities track
observed frequencies closely across the full range (ICI = 0.016).

```{r calibration-plot, fig.width = 5, fig.height = 4}
plot(ad_tone_validation)
```

## Scores by tone category

The Wisconsin Advertising Project (WiscAds) assigned each ad a five-point tone code
ranging from purely promotional to purely attacking. The `textscale` scores align well with these categories, with the score distributions shifting upward as the ads
become more negative.

```{r tone-plot, fig.width = 6.5, fig.height = 4}
ad_tone |>
  filter(!is.na(tone_label)) |>
  ggplot(aes(x = score, y = tone_label)) +
  geom_boxplot() +
  labs(
    x = "textscale Score",
    y = NULL,
    title = "Ad Negativity Scores by Tone Category"
  ) +
  theme_bw()
```

## Correlation with SentimentIt scores

Because the human annotations used to fit the `textscale` model are the same
ones Carlson & Montgomery (2017) used for their SentimentIt model, the two sets of
scores are highly correlated — but they need not be identical, since `textscale`
uses the annotations to learn a direction in embedding space rather than
fitting a SentimentIt model directly.

```{r correlation}
cor(ad_tone$score, ad_tone$alphas, use = "pairwise.complete.obs") |>
  round(3)
```

```{r cm-scatter, fig.width = 6, fig.height = 4.5}
ad_tone |>
  filter(!is.na(tone_label)) |>
  ggplot(aes(x = score, y = alphas, color = tone_label)) +
  geom_point(alpha = 0.7, size = 1.5) +
  labs(
    x     = "textscale Score",
    y     = "SentimentIt Score",
    color = NULL
  ) +
  theme_bw() +
  theme(legend.position = "bottom")
```

## Scoring new documents

Once the model is fit, scoring new documents requires only their embeddings —
no additional pairwise annotations. Here are three hypothetical ads placed on
the negativity scale:

```{r new-docs, eval = FALSE}
new_ads <- c(
  "I will work hard every day in Congress for my constituents. The people
   of Wisconsin are kind, hard-working, and patriotic, and I will do my
   best to emulate their example. Thank you for your support.",

  "Bob McDermott spent over a million dollars in taxpayer funds on a lavish
   vacation in the Bahamas, all while voting to gut Social Security. The
   Wall Street Journal calls him one of the five most corrupt members of
   Congress. Vote to restore sanity to Congress.",

  "Washington is broken, and we need someone from outside the system to fix
   it. I am a plumber with 25 years of experience, so I know a thing or two
   about fixing things. Jim Hokstra has never fixed anything in his life."
)

new_embeddings <- get_embeddings(new_ads)
score_documents(result$model_final, new_embeddings)
```

```
#>      score  lower  upper
#> [1,] -3.20  -3.67  -2.73   # promotional
#> [2,]  3.57   3.07   4.07   # attack
#> [3,]  0.81   0.35   1.27   # contrast
```

The purely promotional ad scores near the bottom of the scale, the attack ad
near the top, and the contrast ad falls in between — consistent with what
the tone categories would predict.

## References

Carlson, T. N., & Montgomery, J. M. (2017). A pairwise comparison framework
for fast, flexible, and reliable human coding of political texts. *American
Political Science Review*, 111(4), 835–843. https://doi.org/10.1017/S0003055417000302
